MSstatsPTM statistical relative quantification of post-translational modifications in global proteomics experiments

### 2022-02-17 main/MSstatsPTM_Main.tex
main intro updates

The scientific community increasingly utilizes mass spectrometry (MS)-based proteomics to detect changes in post-translational modifications (PTMs) between samples from various conditions. Experiments targeting PTMs face numerous challenges. These include the low abundance of modified proteoforms, few representative peptides that span modification sites, and confounding between changes in the abundance of PTM and the overall changes in the protein abundance. Therefore, statistical approaches to estimating relative changes in PTMs, and to detecting differential PTM abundance, should combine information pertaining to PTM sites over several peptides, replicates in multiple conditions, and consider relevant sources of confounding and variation. In this manuscript we propose such a general statistical framework, which is comprehensive, accurate, and leads to reproducible results. The framework requires an experimental design\todo{that, for each sample, quantifies the peptides with post-translational modifications, as well as  peptides from the same proteins with no modification sites}, \todo{and is applicable to both label-free and tandem mass tag (TMT)-based acquisitions}. The proposed framework summarizes the abundances of \todo{peptides with and without the modification sites}, and fits a linear mixed effects model that automatically adjusts to the biological and technological aspects of the experimental design. Finally, changes in the PTM and the protein-level abundances are combined to remove bias due to the confounding of these two sources. Evaluations on computer simulations, a spike-in experiment with known ground truth, and three biological experiments with \todo{which?} acquisition methods \todo{demonstrate the improved - what? - as compared to the currently used approaches for relative PTM quantification}. The proposed framework is implemented in the free and open-source R package $MSstatsPTM$.

@@

The signaling mechanisms that allow cells to mount a dynamic and fast response to a multitude of events are facilitated by the modification of proteins at specific residues, acting as molecular on/off switches\cite{Deribe,Cohen}. Bottom-up liquid chromatography coupled with mass spectrometry (LC-MS) is broadly established as the tool-of-choice for unbiased and large-scale identification and quantification of proteins and their post-translational modifications (PTMs) \cite{Kall:2011ub,Roepstorff}. Studies targeting the post-translationally modified proteome focus either on the accurate localization of modification sites on proteins, relative or absolute quantification of a modification site's occupancy repertoire, or relative changes in occupancy across experimental conditions\cite{Mann}. \todo{The previous sentence is great, but I am wondering whether it can lead the reviewers in a wrong direction, since we only address one of these problems } Some scientific questions,  such as discovery of protein biomarkers of a disease, consider meaningful patterns of changes in PTM abundance alone \cite{Petushkova_2017}. Other questions, e.g. in biological and physiological studies based on very short timescales, require us to distinguish the confounding of quantitative changes in PTM and of overall changes of the protein abundance.

Regardless of the question at hand, interrogating the modified proteome is challenging due to a number of reasons. First, the relatively lower abundance of modified proteoforms dictates that a global interrogation can only be achieved through large-scale enrichment protocols with modification-specific antibodies or beads \todo{refs for these points?}. Variability in the enrichment efficiency inevitably affects the reproducibility of the number of spectral features (e.g., peptide precursor ions or their fragments) and their intensities, which imposes challenges in both quantification and statistical modeling. Second, contrary to the often large number of identified peptides that can be used as features to model protein abundance changes, there are relatively few representative peptides that span a modification sites, which often results in sparse, and sometimes, inherently convoluted models (i.e., single versus multiple modified sites on a single peptide). Third, unless early signaling events are interrogated, the interpretation of the relative changes in modification occupancy are inherently confounded with changes in the overall protein abundance, complication the interpretation of the results \cite{Olsen:2013}. \todo{Finally, technological aspects of bottom-up MS experiments, and presence of labeling, e.g. by tandem mass tag (TMT), introduce additional sources of uncertainty and variation.}

\todo{Currently statistical analyses of these experiments are performed with methods such as t-test and Limma methods\cite{Ritchie_15a}. They have drawbacks XXX}

This manuscript proposes a general statistical analysis framework that detects relative changes in post-translational modifications. The proposed framework combines the quantitative information pertaining to a PTM site over peptides and replicates in multiple conditions, and takes into account various sources of variation and confounding. The framework requires an experimental design\todo{that, for each sample, quantifies the peptides with post-translational modifications, as well as  peptides from the same proteins with no modification sites}. \todo{It is applicable to both label-free and tandem mass tag (TMT)-based acquisitions}. The proposed framework summarizes the abundances of \todo{peptides with and without the modification sites}, and fits a linear mixed effects model that automatically adjusts to the biological and technological aspects of the experimental design. \todo{For questions focusing on PTM, can analyze changes in the modified peptides. For eliminating the confounding, combine...} Finally, changes in the PTM and the protein-level abundances are combined to remove bias due to the confounding of these two sources.

>>>>

The scientific community increasingly utilizes mass spectrometry (MS)-based proteomics to detect changes in post-translational modifications (PTMs) between samples from various conditions. Experiments targeting PTMs face numerous challenges. These include the low abundance of modified proteoforms, few representative peptides that span modification sites, and confounding between changes in the abundance of PTM and the overall changes in the protein abundance. Therefore, statistical approaches to estimating relative changes in PTMs, and to detecting differential PTM abundance, should combine information pertaining to PTM sites over several peptides, replicates in multiple conditions, and consider relevant sources of confounding and variation. In this manuscript we propose such a general statistical framework, which is comprehensive, accurate, and leads to reproducible results. The framework requires an experimental design that, for each sample, quantifies the peptides with post-translational modifications, as well as  peptides from the same proteins with no modification sites, and is applicable to both label-free and tandem mass tag (TMT)-based acquisitions. The proposed framework summarizes the abundances of peptides with and without the modification sites, and fits a linear mixed effects model that automatically adjusts to the biological and technological aspects of the experimental design. Finally, changes in the PTM and the protein-level abundances are combined to remove bias due to the confounding of these two sources. Evaluations on computer simulations, a spike-in experiment with known ground truth, and three biological experiments which demonstrate the improved fold change estimation as compared to the currently used approaches for relative PTM quantification. The proposed framework is implemented in the free and open-source R package $MSstatsPTM$.

@@

The signaling mechanisms that allow cells to mount a dynamic and fast response to a multitude of events are facilitated by the modification of proteins at specific residues, acting as molecular on/off switches\cite{Deribe,Cohen}. Bottom-up liquid chromatography coupled with mass spectrometry (LC-MS) is broadly established as the tool-of-choice for unbiased and large-scale identification and quantification of proteins and their post-translational modifications (PTMs) \cite{Kall:2011ub,Roepstorff}. Studies targeting the post-translationally modified proteome focus either on the accurate localization of modification sites on proteins, relative or absolute quantification of a modification site's occupancy repertoire, or relative changes in occupancy across experimental conditions\cite{Mann}. Some scientific questions,  such as discovery of protein biomarkers of a disease, consider meaningful patterns of changes in PTM abundance alone \cite{Petushkova_2017}. Other questions, e.g. in biological and physiological studies based on very short timescales, require us to distinguish the confounding of quantitative changes in PTM and of overall changes of the protein abundance.

Regardless of the question at hand, interrogating the modified proteome is challenging due to a number of reasons. First, the relatively lower abundance of modified proteoforms dictates that a global interrogation can only be achieved through large-scale enrichment protocols with modification-specific antibodies or beads \cite{Huang:2014}. Variability in the enrichment efficiency inevitably affects the reproducibility of the number of spectral features (e.g., peptide precursor ions or their fragments) and their intensities, which imposes challenges in both quantification and statistical modeling. Second, contrary to the often large number of identified peptides that can be used as features to model protein abundance changes, there are relatively few representative peptides that span a modification sites, which often results in sparse, and sometimes, inherently convoluted models (i.e., single versus multiple modified sites on a single peptide) \cite{Mann}. Third, unless early signaling events are interrogated, the interpretation of the relative changes in modification occupancy are inherently confounded with changes in the overall protein abundance, complication the interpretation of the results \cite{Olsen:2013}. Finally, technological aspects of bottom-up MS experiments, and presence of labeling, e.g. by tandem mass tag (TMT), introduce additional sources of uncertainty and variation.

Currently statistical analyses of these experiments are performed with methods such as t-test and Limma\cite{Ritchie_15a}. These approaches generally use the intensity ratios of modified and unmodified peptide features and compare the mean abundance of different PTM sites. These approaches do not fully account for different sources of variations and as such, are not directly applicable to experiments with complex designs, such as multiple conditions and mixtures. Additionally, there is no self contained implementation of these methods, requiring manual application of the methods by the user and increasing complexity.

This manuscript proposes a general statistical analysis framework that detects relative changes in post-translational modifications. The proposed framework combines the quantitative information pertaining to a PTM site over peptides and replicates in multiple conditions, and takes into account various sources of variation and confounding. The framework requires an experimental design that, for each sample, quantifies the peptides with post-translational modifications, as well as  peptides from the same proteins with no modification sites. It is applicable to both label-free and tandem mass tag (TMT)-based acquisitions. The proposed framework summarizes the abundances of peptides with and without the modification sites, and fits a linear mixed effects model that automatically adjusts to the biological and technological aspects of the experimental design. \todo{For questions focusing on PTM, can analyze changes in the modified peptides. For eliminating the confounding, combine...} Finally, changes in the PTM and the protein-level abundances are combined to remove bias due to the confounding of these two sources.
